---
title: "SET Rate Calculations"
date: "`r Sys.Date()`"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Setup  

The following R packages are required to generate this report:  

```{r load_libraries, echo = TRUE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(here)
```


## Read in data  

This reads in the long dataset, converted from other formats by earlier scripts. It also converts pin heights to mm if they weren't already in those units.    

```{r read_data}
path <- here::here('data/intermediate')
dat <- read_csv(paste0(path, '/GND.csv'))

# make sure pin heights are in mm
if(exists('pin_height_cm', dat)) {
dat <- dat %>%
    mutate(pin_height = pin_height_cm * 10) %>%
    select(-pin_height_cm)
}
if(exists('pin_height_mm', dat)){
    dat <- dat %>%
        mutate(pin_height = pin_height_mm) %>%
        select(-pin_height_mm)
}

# get rid of any placeholders; make sure date is proper format  
dat <- dat %>%
    filter(!is.na(date)) %>%
    mutate(date = as.Date(date))
```

Other reserve-specific inputs here (e.g., NAVD-88 heights of SET receivers; local rate of SLR; etc)  

```{r}
# local rate of sea level rise, in mm/yr
slr <- 3.61
```


***

# Rate Calculations  

## Pin level  

Calculate a linear rate of change for each pin at each SET.  

**Linear regression for each pin.** Keeping errors and confidence intervals in here just so we can describe the linear regression model should it ever be necessary. But really we'll only be working with *rate_mm.yr* from here on out.  

Note that the Confidence Intervals here were generated by the lm() function.  

These values are not shown but can be if people want them.  


```{r, rows.print = 9}
# generate linear model for each pin
models <- dat %>%
    group_by(reserve, set_id, arm_position, pin_number) %>%
    do(mod = lm(pin_height ~ date, data = .))  


# pull out coefficients of interest:
# intercept, slope, p-value, 95% confidence bounds
modelcoef <- tidy(models, mod, conf.int=TRUE, conf.level=0.95) 


# get rid of intercept row,
# calculate mm/yr for all the terms,
# get rid of the mm/day stuff
# p-value and statistic are rounded
rates_elev_pins <- modelcoef %>%
    filter(term == "date") %>%
    mutate(rate_mm.yr = estimate*365.25,
           se_mm.yr = std.error*365.25,
           CIlow_mm.yr = conf.low*365.25,
           CIhigh_mm.yr = conf.high*365.25,
           p_value = round(p.value, 4),
           statistic = round(statistic, 3)) %>%
    select(reserve, set_id, arm_position, pin_number, rate_mm.yr, se_mm.yr, CIlow_mm.yr, CIhigh_mm.yr, statistic, p_value)


# calculate r^2, adjusted r^2, p values, and other model diagnostics
modelsummary_pins <- glance(models, mod) 
```

## Arm Level  

The average of 9 pins in an arm. Again, values are not shown, but can be.  

```{r}
rates_elev_arms <- rates_elev_pins %>%
    group_by(reserve, set_id, arm_position) %>%
    summarize(rate_mm.yr_arm = mean(rate_mm.yr, na.rm = TRUE),
              sd_mm.yr = sd(rate_mm.yr, na.rm = TRUE),
              se_mm.yr = sd_mm.yr/sqrt(n())) %>%
    ungroup()
```

***


## SET level: All rates  

Average of the 4 arms.  

95% Confidence Intervals are calculated as:  
`mean +/- [1.96 * standard error]`  

<br>

```{r}
rates_elev_set <- rates_elev_arms %>%
    group_by(reserve, set_id) %>%
    summarize(rate_mm.yr_set = mean(rate_mm.yr_arm, na.rm = TRUE),
              sd_mm.yr = sd(rate_mm.yr_arm, na.rm = TRUE),
              se_mm.yr = sd_mm.yr/sqrt(n())) %>%
    mutate(CI95_low = rate_mm.yr_set - 1.96*se_mm.yr,
           CI95_high = rate_mm.yr_set + 1.96*se_mm.yr) %>%
    ungroup()

knitr::kable(rates_elev_set,
             caption = "Rate of Elevation Change by SET", 
             col.names = c("Reserve", "SET ID", "Rate (mm/yr)", "stdev", "sterr", "95% CI lower bound", "95% CI upper bound"),
             digits = 3,
             align = 'c')
```

***


# Sea Level Rise Comparisons  

The local rate of sea level rise was input as **`r slr` mm/yr** .

The following tables break the SETs into groups where the rate of SET elevation change is *lower than* / *higher than* / *not different from* this SLR rate.  

95% Confidence Intervals are calculated as:  
`mean +/- [1.96 * standard error]`  

<br>

***

<br>

```{r}
rates_elev_set %>%
    filter(CI95_high < slr) %>%
    select(reserve, set_id, rate_mm.yr_set, CI95_low, CI95_high) %>%
    knitr::kable(caption = "SETs where elevation change is lower than SLR",
                 col.names = c("Reserve", "SET ID", "Rate (mm/yr)", "95% CI lower bound", "95% CI upper bound"),
                 digits = 3,
                 align = 'c')
```

<br>

***

<br>

```{r}
rates_elev_set %>%
    filter(CI95_low > slr) %>%
    select(reserve, set_id, rate_mm.yr_set, CI95_low, CI95_high) %>%
    knitr::kable(caption = "SETs where elevation change is higher than SLR",
                 col.names = c("Reserve", "SET ID", "Rate (mm/yr)", "95% CI lower bound", "95% CI upper bound"),
                 digits = 3,
                 align = 'c')
```

<br>

***

<br>

```{r}
rates_elev_set %>%
    filter(CI95_low < slr & CI95_high > slr) %>%
    select(reserve, set_id, rate_mm.yr_set, CI95_low, CI95_high) %>%
    knitr::kable(caption = "SETs where the 95% CI for elevation change rate includes SLR",
                 col.names = c("Reserve", "SET ID", "Rate (mm/yr)", "95% CI lower bound", "95% CI upper bound"),
                 digits = 3,
                 align = 'c')
```

***  

# Graph  

for a sanity check of the above rates  

Linear regression is NOT appropriate for all sites; interpret this with caution

```{r, fig.width = 7, fig.height = 10}
ggplot(dat, aes(x = date, y = pin_height)) +
    geom_point(aes(col = arm_position), alpha = 0.6) +
    facet_wrap(~set_id, ncol = 4, scales = "free_y") +
    geom_smooth(method = 'lm', se = FALSE, size = 2, col = "black") +
    labs(title = "Pin readings by date", subtitle = "with linear regression through all pins", x = "Date", y = "Pin Height (mm)") +
    scale_fill_discrete(name = 'Arm Position') +
    theme_bw() +
    theme(legend.position = 'bottom')
```


```{r}
# For statistical comparison purposes. What happens if we average all 36 pins' rates up to the SET level, skipping the average-to-arm step? 


# rates_elev_set2 <- rates_elev_pins %>%
#     group_by(reserve, set_id) %>%
#     summarize(rate_mm.yr_set = mean(rate_mm.yr, na.rm = TRUE),
#               sd_mm.yr = sd(rate_mm.yr, na.rm = TRUE),
#               se_mm.yr = sd_mm.yr/sqrt(n())) %>%
#     mutate(CI95_low = rate_mm.yr_set - 1.96*se_mm.yr,
#            CI95_high = rate_mm.yr_set + 1.96*se_mm.yr) %>%
#     ungroup()
# 
# knitr::kable(rates_elev_set2,
#              caption = "SET change rates calculated from 36 pins rather than 4 arms",
#              col.names = c("Reserve", "SET ID", "Rate (mm/yr)", "stdev", "sterr", "95% CI lower bound", "95% CI upper bound"),
#              digits = 3,
#              align = 'c')
```



